%!TEX root =  main.tex

\chapter{Connection to Gaussian-Process Regression}

\ac{GP} regression is a popular method for function approximation that fits with the Bayesian estimation paradigm \citep{rasmussen06}.  In this chapter, we will show that the continuous-time estimation framework we have developed in the previous chapter is identical to \ac{GP} regression.  This means that the factor-graph framework we have developed is also a powerful tool for \ac{GP} regression (with our particular choice of kernels).  We primarily follow \citet{barfoot_ser24} in this chapter.

\section{Brief Derivation of GP Regression}

In general, we can consider systems with a continuous-time \ac{GP} prior and a discrete-time, linear measurement model:
\beqn{}
\mbf{x}(t) & \sim & \mathcal{GP}( \pri{\mbf{x}}(t), \pri{\mbf{P}}(t,t^\prime) ), \qquad t_0 < t,t^\prime, \\
\mbf{y}_k & = & \mbf{C}_k\mbf{x}(t_k) + \mbf{n}_k, \qquad t_0 < t_1 < \cdots < t_K,
\eeqn 
where $\mbf{x}(t)$ is the state, $\pri{\mbf{x}}(t)$ is the mean function, $\pri{\mbf{P}}(t,t^\prime)$ is the covariance function, $\mbf{y}_k$ are measurements, $\mbf{n}_k \sim \mathcal{N}(\mbf{0}, \mbf{R}_k )$ is Gaussian measurement noise, and $\mbf{C}_k$ is the measurement model coefficient matrix. Initially, $\mbf{x}(0) = \mathcal{N}( \pri{\mbf{x}}_0, \pri{\mbf{P}}_0 )$. The first equation is a {\em kernel function} that describes the covariance between the state at two different times, $t$ and $t^\prime$; it captures our prior belief about the likelihood of all possible trajectories that the state could take. 

We consider that we want to query the state at a number of times ($\tau_0 < \tau_1< \ldots < \tau_J$) that may or may not be different from the measurement times ($t_0 < t_1 < \ldots < t_K$). 
The joint density between the state (at the query times) and the measurements (at the measurement times) is written as
\begin{equation}\label{eq:gaussjoint}
p\left( \bbm   \mbf{x}_\tau \\ \mbf{y} \ebm \right) = \mathcal{N} \left( \bbm   \pri{\mbf{x}}_\tau  \\ \mbf{C} \pri{\mbf{x}} \ebm, \bbm  \pri{\mbf{P}}_{\tau\tau} & \pri{\mbf{P}}_\tau \mbf{C}^T \\ \mbf{C} \pri{\mbf{P}}_\tau^T &   \mbf{R} + \mbf{C} \pri{\mbf{P}}\mbf{C}^T \ebm  \right),
\end{equation}
where
\begin{gather*}
\mbf{x} = \bbm \mbf{x}(t_0) \\ \vdots \\ \mbf{x}(t_K) \ebm,  \quad \pri{\mbf{x}} = \bbm \pri{\mbf{x}}(t_0)  \\ \vdots \\ \pri{\mbf{x}}(t_K) \ebm, \quad \mbf{x}_\tau = \bbm \mbf{x}(\tau_0) \\ \vdots \\ \mbf{x}(\tau_J) \ebm,  \quad \pri{\mbf{x}}_\tau = \bbm \pri{\mbf{x}}(\tau_0) \\ \vdots \\ \pri{\mbf{x}}(\tau_J) \ebm, \\
\mbf{y} = \bbm \mbf{y}_0 \\ \vdots \\ \mbf{y}_K \ebm,  \quad \mbf{C} = \mbox{diag}\left( \mbf{C}_0, \ldots, \mbf{C}_K \right), \quad \mbf{R} = \mbox{diag}\left( \mbf{R}_0, \ldots, \mbf{R}_K \right), \\
\pri{\mbf{P}} = \bbm \pri{\mbf{P}}(t_i, t_j) \ebm_{ij}, \quad \pri{\mbf{P}}_\tau = \bbm \pri{\mbf{P}}(\tau_i, t_j) \ebm_{ij}, \quad \pri{\mbf{P}}_{\tau\tau} = \bbm \pri{\mbf{P}}(\tau_i, \tau_j) \ebm_{ij}.
\end{gather*}
In \ac{GP} regression, the matrix, $\pri{\mbf{P}}$, is known as the \index{kernel matrix} {\em kernel matrix} and comprises evaluations of the kernel covariance function between all pairs of measurement times.  Factoring~\eqref{eq:gaussjoint}, we can show that
\begin{multline}
\label{eq:gpquerytimes}
p(\mbf{x}_\tau | \mbf{y}) = \mathcal{N}\biggl( \underbrace{\pri{\mbf{x}}_\tau + \pri{\mbf{P}}_{\tau} \mbf{C}^T( \mbf{C} \pri{\mbf{P}} \mbf{C}^T + \mbf{R})^{-1} (\mbf{y} - \mbf{C}\pri{\mbf{x}} )}_{\est{\mbf{x}}_{\tau}, \;\mbox{\small mean}}, \biggr. \\ \biggl. 
\underbrace{\pri{\mbf{P}}_{\tau\tau} - \pri{\mbf{P}}_\tau \mbf{C}^T ( \mbf{C} \pri{\mbf{P}} \mbf{C}^T + \mbf{R}  )^{-1} \mbf{C} \pri{\mbf{P}}_\tau^T }_{\est{\mbf{P}}_{\tau\tau}, \;\mbox{\small covariance}}\biggr),
\end{multline}
for the density of the predicted state at the query times, given the measurements.  

The expression simplifies further if we take the query times to be exactly the same as the measurement times (i.e., $\tau_k = t_k, K=J$).  This implies that 
\begin{equation}
\pri{\mbf{P}} = \pri{\mbf{P}}_\tau = \pri{\mbf{P}}_{\tau\tau},
\end{equation}
and then we can write
\begin{multline}
\label{eq:gpmeastimes2}
p(\mbf{x} | \mbf{y}) = \mathcal{N}\biggl( \underbrace{\pri{\mbf{x}} + \pri{\mbf{P}} \mbf{C}^T( \mbf{C} \pri{\mbf{P}} \mbf{C}^T + \mbf{R})^{-1} (\mbf{y} - \mbf{C}\pri{\mbf{x}} )}_{\est{\mbf{x}}, \;\mbox{\small mean}}, \biggr. \\ \biggl.
\underbrace{\pri{\mbf{P}} - \pri{\mbf{P}} \mbf{C}^T ( \mbf{C} \pri{\mbf{P}} \mbf{C}^T + \mbf{R}  )^{-1} \mbf{C} \pri{\mbf{P}}^T }_{\est{\mbf{P}}, \;\mbox{\small covariance}}\biggr).
\end{multline}
We can rearrange this using the {\em matrix-inversion lemma} so that
\begin{multline}
\label{eq:gpmeastimes}
p(\mbf{x} | \mbf{y}) = \mathcal{N}\biggl(  \underbrace{\left( \pri{\mbf{P}}^{-1} + \mbf{C}^T \mbf{R}^{-1} \mbf{C} \right)^{-1} \left( \pri{\mbf{P}}^{-1} \pri{\mbf{x}}  + \mbf{C}^T \mbf{R}^{-1}\mbf{y} \right)}_{\est{\mbf{x}}, \;\mbox{\small mean}}, \biggr. \\ \biggl.\underbrace{\left( \pri{\mbf{P}}^{-1} + \mbf{C}^T \mbf{R}^{-1} \mbf{C} \right)^{-1} }_{\est{\mbf{P}}, \;\mbox{\small covariance}} \biggr).
\end{multline}
Further rearranging the mean expression, we have a linear system for $\est{\mbf{x}}$:
\begin{equation}
\label{eq:gpgnlin}
 \left( \pri{\mbf{P}}^{-1} + \mbf{C}^T \mbf{R}^{-1} \mbf{C} \right) \, \est{\mbf{x}} =  \pri{\mbf{P}}^{-1} \pri{\mbf{x}} + \mbf{C}^T \mbf{R}^{-1}\mbf{y}.
\end{equation}
The solution to this linear system gives us the estimate of the state at the measurement times.

If, after solving for the estimate at the measurement times, we later want to query the state at some other times of interest ($\tau_0 < \tau_1< \ldots < \tau_J$), we can use the classic \ac{GP}  interpolation equations \citep{rasmussen06} to do so:
\begin{subequations}
\label{eq:gpinterp}
\begin{eqnarray}
\est{\mbf{x}}_\tau & = & \pri{\mbf{x}}_\tau + \left( \pri{\mbf{P}}_\tau \pri{\mbf{P}}^{-1} \right) (\est{\mbf{x}} - \pri{\mbf{x}}), \\
\est{\mbf{P}}_{\tau\tau} & = & \pri{\mbf{P}}_{\tau\tau} + \left( \pri{\mbf{P}}_\tau \pri{\mbf{P}}^{-1}\right) \left( \est{\mbf{P}} - \pri{\mbf{P}} \right) \left( \pri{\mbf{P}}_\tau \pri{\mbf{P}}^{-1}\right)^T.
\end{eqnarray}
\end{subequations}
This is linear interpolation in the state variable (but not necessarily in time).  

In general, this \index{Gaussian process} \ac{GP} approach has complexity $O(K^3 + K^2J)$ since the initial solve is $O(K^3)$ (for an arbitrary kernel function such as squared exponential) and the query is $O(K^2J)$; this is quite expensive, and next we will seek to improve the cost by designing a special kernel function.

\section{Exactly Sparse Kernels}

Looking at~\eqref{eq:gpgnlin}, we see that the {\em inverse kernel matrix}, $\pri{\mbf{P}}^{-1}$, appears on the left-hand side.  Naively choosing a \ac{GP} kernel function could result in this matrix being dense and therefore the solution expensive.  Instead, we will select our kernel function carefully so that the inverse kernel matrix is {\em exactly sparse}.

We base our kernel prior on a \ac{LTI}\footnote{We could also make this equation time-varying \citep{barfoot_ser24} but leave it time-invariant in the interest of simplicity.} \acf{SDE}:
\begin{equation}
\dot{\mbs{x}}(t) = \mbs{A} \mbs{x}(t) + \mbs{v}(t) + \mbs{L} \mbs{w}(t),
\end{equation}
with
\begin{equation}
\mbs{w}(t) \sim \mathcal{GP}(\mbf{0}, \mbs{Q} \, \delta(t-t^\prime)),
\end{equation}
a (stationary) zero-mean \index{Gaussian process} \ac{GP} with (symmetric, positive-definite) \index{power spectral density matrix} {\em power spectral density matrix}, $\mbs{Q}$.  This is the same motion model that we introduced in~\eqref{eq:linsysct}.

Referring to \citep{barfoot_ser24}, we can show that the solution to this \ac{SDE} is a \ac{GP} with the following kernel function:
\begin{multline}
\mbs{x}(t) \sim \mathcal{GP} \biggl(  \underbrace{\mbs{\Phi}(t,t_0)\pri{\mbf{x}}_0 + \int_{t_0}^t \mbs{\Phi}(t,s) \mbs{v}(s) \, ds}_{\pri{\mbf{x}}(t)},  \\ \underbrace{\mbs{\Phi}(t,t_0) \pri{\mbf{P}}_0 \mbs{\Phi}(t^\prime,t_0)^T + \int_{t_0}^{\min(t,t^\prime)} \mbs{\Phi}(t,s) \mbs{L}\mbs{Q}\mbs{L}^T \mbs{\Phi}(t^\prime,s)^T  \, ds}_{\pri{\mbf{P}}(t,t^\prime)} \biggr),
\end{multline}
where $\mbs{\Phi}(t,s)$ is known as the {\em transition function}.  In our case, the system is \ac{LTI} and so the transition function is given by
\begin{equation}
\mbs{\Phi}(t,s) = \exp\left( \mbs{A}(t-s) \right).
\end{equation}
Evaluating this kernel function at the measurement times allows us to write our prior as a joint Gaussian density:
\begin{equation}
\label{eq:gpprior}
p(\mbf{x}) = \mathcal{N}(\pri{\mbf{x}}, \pri{\mbf{P}}) = \mathcal{N} \left( \mbf{A} \mbf{v}, \mbf{A} \mbf{Q} \mbf{A}^T \right),
\end{equation}
where
\begin{gather}
\mbf{v} = \bbm \pri{\mbf{x}}_0 \\ \mbf{v}_{1,0} \\ \vdots \\ \mbf{v}_{K,K-1} \ebm, \quad 
\mbf{Q} = \diag\left( \pri{\mbf{P}}_0, \mbf{Q}_{1,0}, \mbf{Q}_{2,1}, \ldots, \mbf{Q}_{K,K-1} \right), \\
\mbf{v}_{k,k-1} = \int_{t_{k-1}}^{t_k} \mbs{\Phi}(t_k,s) \mbs{v}(s) \, ds, \\
\mbf{Q}_{k,k-1} = \int_{t_{k-1}}^{t_k} \mbs{\Phi}(t_k,s) \mbs{L} \mbs{Q} \mbs{L}^T \mbs{\Phi}(t_k,s)^T \, ds, \nonumber \\
 \mbf{A} = \bbm \mbf{1} &  &  &  &   & \\
\mbf{A}_{1,0} & \mbf{1} &  &  &  &  \\
\mbf{A}_{2,0} & \mbf{A}_{2,1} & \mbf{1} & &  &  \\
\vdots & \vdots & \vdots & \ddots &  & \\
\mbf{A}_{K-1,0} & \mbf{A}_{K-1,1} & \mbf{A}_{K-1,2} & \cdots & \mbf{1} &  \\
\mbf{A}_{K,0} & \mbf{A}_{K,1} & \mbf{A}_{K,2} & \cdots & \mbf{A}_{K,K-1} & \mbf{1}
\ebm, \nonumber
\end{gather}
and we have simplified notation by writing $\mbf{A}_{k,\ell} = \mbs{\Phi}(t_k, t_\ell)$.

Substituting our specific prior into~\eqref{eq:gpgnlin}, the linear system that we need to solve becomes
\begin{equation}
\label{eq:gpgnlinsparse}
\left( \mbf{A}^{-T} \mbf{Q}^{-1} \mbf{A}^{-1} + \mbf{C}^T \mbf{R}^{-1} \mbf{C} \right) \, \est{\mbf{x}} =  \mbf{A}^{-T} \mbf{Q}^{-1} \mbf{v} + \mbf{C}^T \mbf{R}^{-1}\mbf{y},
\end{equation}
which as yet does not seem to offer any advantage over~\eqref{eq:gpgnlin}.  However, it turns out that $\mbf{A}^{-1}$ has the special form
\begin{equation}
\mbf{A}^{-1} = \bbm \mbf{1} &  & &  &  &  \\
-\mbf{A}_{1,0} & \mbf{1} &  &  &  &  \\
& -\mbf{A}_{2,1} & \mbf{1} & &  &  \\
 &  & -\mbf{A}_{3,2} & \ddots &  &  \\
 &  & & \ddots & \mbf{1} &  \\
 &  &  &  & -\mbf{A}_{K,K-1} & \mbf{1}
\ebm.
\end{equation}
Since $\mbf{A}^{-1}$ has only the main diagonal and the one below non-zero, and $\mbf{Q}^{-1}$ is block-diagonal, the inverse kernel matrix $\pri{\mbf{P}}^{-1} = (\mbf{A}\mbf{Q}\mbf{A}^T)^{-1} = \mbf{A}^{-T} \mbf{Q}^{-1} \mbf{A}^{-1}$ is exactly block-tridiagonal.  Since $\mbf{C}^T \mbf{R}^{-1} \mbf{C}$ is block-diagonal, it means the left-hand side of~\eqref{eq:gpgnlinsparse} is also block-tridiagonal.  This means that we can solve the linear system in $O(K)$ time instead of $O(K^3)$ for a dense inverse kernel matrix.  This has been made possible by our specific choice of kernel function, built from an \ac{LTI} \ac{SDE}.

We can also view~\eqref{eq:gpgnlinsparse} as the solution to the following least-squares optimization problem:
\begin{equation}
\est{\mbf{x}} = \mbox{arg}\min_{\mbf{x}} || \mbf{v} - \mbf{A}^{-1}\,\mbf{x} ||^2_{\mbf{Q}} + || \mbf{y} - \mbf{C}\mbf{x}||^2_{\mbf{R}}.
\end{equation}
Or, breaking it down by time steps, 
\begin{multline}
\est{\mbf{x}} = \mbox{arg}\min_{\mbf{x}} || \pri{\mbf{x}}_0 - \mbf{x}_0 ||^2_{\pri{\mbf{P}}_0} + \sum_{k=1}^K || \mbf{x}_k - \mbf{A}_{k,k-1} \mbf{x}_{k-1} - \mbf{v}_{k,k-1} ||^2_{\mbf{Q}_{k,k-1}} \\ + \sum_{k=0}^K || \mbf{y}_k - \mbf{C}_k \mbf{x}_k||^2_{\mbf{R}_k}.
\end{multline}
Each of these terms accounts for exactly one factor in Figure~\ref{fig:dtfg2}.  Therefore, solving at the measurement times is equivalent to our earlier factor-graph formulation for the main solve.

\section{Querying is $O(1)$}

Next, we will show that querying the trajectory at any time $\tau$ is an $O(1)$ operation.  To do this, we will use the interpolation equations~\eqref{eq:gpinterp}, where we notice that the product $\pri{\mbf{P}}_\tau \pri{\mbf{P}}^{-1}$ plays a critical role.  To simplify matters, let us assume that we have only a single query time, $t_{k-1} < \tau < t_k$.  Then we can show that
\begin{multline}
\pri{\mbf{P}}_\tau \pri{\mbf{P}}^{-1} = \Bigl[ \, \mbf{0} \;\;\; \cdots \;\;\;  \mbf{0} \;\;\; \underbrace{\mbf{A}_{\tau,k-1} - \mbf{Q}_{\tau,k-1} \mbf{A}_{k,\tau}^T \mbf{Q}_{k,k-1}^{-1} \mbf{A}_{k,k-1}}_{\mbs{\Lambda}_\tau\; \mbox{\small block column $k-1$}}  \\  \cdots \;\;\;  \underbrace{\mbf{Q}_{\tau,k} \mbf{A}_{k,\tau}^T \mbf{Q}_{k,k-1}^{-1}}_{\mbs{\Psi}_\tau,\;\mbox{\small block column $k$}}  \;\;\; \mbf{0} \;\;\; \cdots \;\;\; \mbf{0}  \, \Bigr],
\end{multline}
so that only block-columns $k-1$ and $k$ are non-zero.  This means that we only require the marginal associated with the joint density of $\mbf{x}_{k-1}$ and $\mbf{x}_k$ to compute the query at time $\tau$:
\beqn{}
\est{\mbf{x}}_\tau & = & \pri{\mbf{x}}_\tau + \bbm \mbs{\Lambda}_\tau & \mbs{\Psi}_\tau \ebm \left( \bbm \est{\mbf{x}}_{k-1} \\ \est{\mbf{x}}_{1} \ebm - \bbm \pri{\mbf{x}}_{k-1} \\  \pri{\mbf{x}}_{k} \ebm \right), \label{eq:meaninterp} \\
\est{\mbf{P}}_\tau & = & \pri{\mbf{P}}_\tau + \bbm \mbs{\Lambda}_\tau & \mbs{\Psi}_\tau \ebm \left( \bbm \est{\mbf{P}}_{k-1,k-1} & \est{\mbf{P}}_{k-1,k}^T \\ \est{\mbf{P}}_{k,k-1} & \est{\mbf{P}}_{k,k} \ebm \right.  \label{eq:covinterp} \\ & & \qquad\qquad\qquad\qquad \quad\left. - \bbm \pri{\mbf{P}}_{k-1,k-1} & \pri{\mbf{P}}_{k,k-1}^T \\ \pri{\mbf{P}}_{k,k-1} & \pri{\mbf{P}}_{k,k} \ebm \right) \bbm \mbs{\Lambda}_\tau^T \\ \mbs{\Psi}_\tau^T \ebm. \nonumber
\eeqn
It turns out this is identical to~\eqref{eq:fgquery}, which requires that we show
\beqn{}
\mbs{\eta}_t & = & \pri{\mbf{x}}_\tau - \bbm \mbs{\Lambda}_\tau & \mbs{\Psi}_\tau \ebm \bbm \pri{\mbf{x}}_{k-1} \\  \pri{\mbf{x}}_{k} \ebm, \\
\mbs{\Sigma}_t & = & \pri{\mbf{P}}_\tau - \bbm \mbs{\Lambda}_\tau & \mbs{\Psi}_\tau \ebm \bbm \pri{\mbf{P}}_{k-1,k-1} & \pri{\mbf{P}}_{k,k-1}^T \\ \pri{\mbf{P}}_{k,k-1} & \pri{\mbf{P}}_{k,k} \ebm \bbm \mbs{\Lambda}_\tau^T \\ \mbs{\Psi}_\tau^T \ebm,
\eeqn
left an exercise for the reader.  

This means that using the traditional \ac{GP} interpolation equations is exactly equivalent to our factor-graph elimination approach from the previous chapter.  We argue that the factor-graph approach is more intuitive and elegant because the sparse structure of the problem is plain to see; the separator for $\mbf{x}_\tau$ is clearly $\{ \mbf{x}_{k-1}, \mbf{x}_k\}$, so that the query will clearly be an $O(1)$ operation. 

